### 2021-07-03 chaps/thesis/chap1/sections/article.tex
📝 use physics package

The goal of this work is to develop high order and efficient numerical methods for nonlinear \textcolor{black}{collisionless} kinetic models, such as the Vlasov-Poisson equations or drift-kinetic models. In most situations, the nonlinearity in the transport term originates from the coupling with a Poisson type problem that is used to compute the electric field.

@@

\section{Exponential integrators and Lawson methods \label{sec:expint}}

@@

In this paper, we will consider two types of exponential methods. The idea of \textit{exponential integrators} is to use the variation of constants formula to rewrite equation \eqref{eq:expint-eq} in the following form
\[ u(t_n+\Delta t) = \exp(\Delta t A) u(t_n) + \int_0^{\Delta t} \exp((\Delta t -s)A) F(u(t_n+s)) \,\mathrm{d}s, \]
where we denote the time step size by $\Delta t >0$ and $t_n = n\Delta t$ with $n\in\mathbb{N}$. This expression is still exact; \ie no approximation has been made. Note, however, that this can not be used as a numerical method as evaluating the integral would require the knowledge of $u(t_n+s)$, which is not available. \textcolor{black}{The idea of an exponential integrator is to approximate the nonlinear part $F(u(t_n+s))$ in terms of the available data. In the simplest case we just evaluate it at the left endpoint. That is, we use $F(u(t_n+s))\approx F(u^n)$. Then we can integrate the term $\exp((\Delta t -s)A)$ exactly and obtain.}
\[ u(t_n+\Delta t) \approx u^{n+1} = \exp(\Delta t A) u^n + \Delta t \varphi_1(\Delta t A) F(u^n), \]
where $\varphi_1(z)=(\mathrm{e}^z-1)/z$ is an entire function. This is the first order exponential Euler method. In a similar way exponential Runge--Kutta methods can be constructed. We refer to the literature, in particular the review article \cite{Hochbruck:2010}, for more details.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Linear analysis \label{ode}}

@@

\subsection{Lawson methods}

@@

\subsection{Exponential integrators}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Phase space discretization}

@@

\paragraph{Centered scheme in $v$.\\} The classic centered scheme is obtained by approximating the velocity derivative in equation \eqref{fourier_x_vlasov} by
$$
(\partial_v \hat{f}_{k})(v_j) \approx \frac{\hat{f}_{k, j+1}-\hat{f}_{k, j-1}}{2\Delta v},
$$ 
where $\hat{f}_{k,j}$ is an approximation of $\hat{f}_{k}(v_j)$.  Inserting this centered approximation 
in \eqref{fourier_x_vlasov} yields
\begin{equation} \label{eq:half-fourier}
\partial_t \hat{f}_{k,j} +  i d k\hat{f}_{k,j} +b \frac{\hat{f}_{k,j+1} -\hat{f}_{k,j-1} }{2\Delta v} = 0. \end{equation}
The system is already diagonal with respect to the index $k$. We now also diagonalize it with respect to the index $j$. To do that we express the function in terms of its Fourier modes with respect to $v$. That is,
\[ \hat{f}_{k,j} = \sum_m \bar{f}_{k, m}\exp\left(i \frac{2\pi m}{2v_{\max}}  v_j\right), \]
where $\bar{f}_{k,m}$ denotes the (double) Fourier transform of $f$ with frequency in space $k$ and frequency in velocity $m$. Inserting this into equation \eqref{eq:half-fourier} yields
\begin{equation}
\label{discrete_linear_transport}
	\partial_t \bar{f}_{k,m} + i dk\bar{f}_{k,m} +b \frac{i\sin(2\pi m \Delta v/(2v_{\max})) }{\Delta v} \bar{f}_{k,m}= 0.  
\end{equation}
We immediately see that this equation is precisely in the form of equation \eqref{ode_linear} as studied in the previous section. We also observe that $\lambda \in i\mathbb{R}$. That is, the eigenvalues for the centered difference approximation lie exclusively on the imaginary axis. One immediate consequence is that for Lawson methods the CFL condition is given by $b \Delta t < C \Delta v$, where $C$ is chosen such that $i [-C,C]$ lies in the domain of stability of the underlying Runge--Kutta method.

\paragraph{Linearized WENO approximation in $v$.\\}

A common technique to discretize hyperbolic partial differential equations is to use the so-called weighted essentially non-oscillatory schemes (WENO) schemes. These are nonlinear schemes that limit oscillations in regions where sharp gradients occur, but still yield high order accuracy in smooth regions of the phase space. In the linear case WENO schemes reduce to upwind discretizations. Here, we will consider the LW5 scheme (the linearized version of the WENO5 scheme as considered in \cite{Baldauf:2008, Lunet:2017, Motamed:2010, Wang:2007}) that is given by (from now on we assume w.l.o.g.~that $b>0$)
\begin{align*}
(\partial_v \hat{f}_k)(v_j) &\approx \frac{1}{\Delta v}\Big(-\frac{1}{30} \hat{f}_{k,j-3} +\frac{1}{4} \hat{f}_{k,j-2} -\hat{f}_{k,j-1} + \frac{1}{3} \hat{f}_{k,j} +\frac{1}{2} \hat{f}_{k,j+1} - \frac{1}{20} \hat{f}_{k,j+2}\Big). 
\end{align*}
We now perform the same analysis as for the centered scheme (see \cite{Baldauf:2008, Crouseilles:2012}). This yields
\begin{align}
  {\color{black} \left( \frac{2\pi }{2 v_{\max}}  i m \bar{f}_{k, m} \approx \right)}
   %(i m \bar{f}_{k, m} \approx ) \;\; 
   \mu_m \bar{f}_{k, m} &:= \frac{\bar{f}_{k,m}}{\Delta v}\Big(-\frac{1}{30} e^{-\frac{3i m\pi \Delta v}{v_{\max}}} +\frac{1}{4} e^{-\frac{2i m \pi \Delta v}{v_{\max}}} -e^{-\frac{i m \pi \Delta v}{v_{\max}}}  \nonumber\\
&\hspace{1cm}+ \frac{1}{3}  +\frac{1}{2} e^{\frac{i m \pi \Delta v}{v_{\max}}} - \frac{1}{20} e^{\frac{2i m \pi \Delta v}{v_{\max}}}\Big). 
\label{lw5symbol}
\end{align}
We then obtain the equation 
\begin{equation}
	\label{discrete_linear_transport_weno}
	\partial_t \bar{f}_{k,m} + i dk\bar{f}_{k,m} +b \mu_m \bar{f}_{k,m}= 0.
\end{equation}
Once again this is precisely the form of equation \eqref{ode_linear}, where $a=dk\in\mathbb{R}$ and $\lambda=b\mu_m \in\mathbb{C}$. The main difference to the centered difference scheme is that $\lambda$ is not necessarily on the imaginary axis. In fact, the eigenvalues aquire a negative real part which stabilizes the scheme and avoids spurious oscillations, but also adds unphysical dissipation to the numerical method.

{\color{black} 
\begin{remark}
Let us remark that performing a Fourier approximation in $v$ is also possible. Using the same notation as before, the counterpart of \eqref{discrete_linear_transport} and \eqref{discrete_linear_transport_weno} in that case is
$$
\partial_t \bar{f}_{k,m} + i dk\bar{f}_{k,m} +b i \frac{2\pi m}{2 v_{\max}} \bar{f}_{k,m}= 0.
$$
It is worth mentioning that this last equation can be obtained by considering the limit $\Delta v$ goes to zero in  \eqref{discrete_linear_transport} or \eqref{discrete_linear_transport_weno}. The stability condition can be computed as for the centered difference case, since the eigenvalues of the Fourier approximation also lies on the imaginary axis. The CFL condition is given by $b \pi  < C \Delta v$, where $C$ is chosen such that $i [-C,C]$ lies in the domain of stability of the underlying Runge--Kutta method.

\end{remark}
}

\subsection{Computing the CFL condition}

@@

\subsubsection{Centered scheme in $v$.}
In the case of centered approximation of the velocity derivative, the Fourier multiplier is a 
pure imaginary complex number (see equation \eqref{discrete_linear_transport}). We thus look for $y_{\max}\in \mathbb{R}_+$ such that the interval $i(-y_{\max},y_{\max}) \subset {\cal D}$, where ${\cal D}$ is the domain of stability for the chosen time integrator.

\paragraph{Lawson integrators.\\} 
We simply look for the largest value $y_{\max}$ such that $i(-y_{\max}, y_{\max}) \subset \mathcal{D}$. The corresponding values for a number of schemes are listed in Table \ref{tab:ymax_Lawson}. These values have to be understood in the 
following way: they induce the CFL condition $b \Delta t\leq y_{\max}\Delta v$ for the discretized equation \eqref{discrete_linear_transport}, where $\Delta t$ denotes the time step size and $\Delta v$ is the velocity mesh size.

@@

\paragraph{Exponential integrators. \\}

For the exponential integrators the domain of stability is very sensitive to the value of $(a\Delta t)$. To get an idea of what we can expect, we consider the quantity $$y_{\max}=\min_{(a\Delta t)\in\mathbb{R}} \; y^{exp}_{\max}(a \Delta t).$$ As before, $y^{exp}_{\max}(a\Delta t)$ is the largest value such that : $$i(-y^{exp}_{\max}(a\Delta t), y^{exp}_{\max}(a\Delta t))\subset {\cal D},$$ where ${\cal D}$ is the domain of stability for the chosen exponential time integrator for a given $(a\Delta t)$. 
Even for relatively simple numerical methods it is not possible to compute this quantity analytically. Thus, we resort to numerical approximations. Unfortunately, it turns out that for most exponential integrators this value is zero. This can be appreciated by considering Figure \ref{fig:expRK_sd} once more. Clearly, there are values of $(a \Delta t)$ such that no relevant part of the imaginary axis (or only half the imaginary axis) is part of the domain of stability.
Thus, most exponential integrators are unstable in the von Neumann sense. However, this is not what we observe in practice. In fact, already the results presented in \cite{Crouseilles:2018} indicate that we can successfully run numerical simulations using, for example, the Cox--Matthews scheme. There are two major points to consider here
\begin{itemize}
    \item The $y_{\max}$ obtained is a worst case estimate. In fact, we know that for $\Delta t \to 0$ we regain the stability of the underlying Runge--Kutta method. Thus, for small $(a \Delta t)$ the methods is expected to work well.
    \item As is usually done we have mandated that $\vert \phi(z) \vert \leq 1$. However, strictly speaking this is not necessary for practical simulation. If we assume that $\vert \phi(z) \vert \leq 1+\varepsilon$ and we take $n$ steps the amplification of the error is given by $(1+\varepsilon)^n$. In the limit $n \to +\infty$ this quantity diverges. However, since we usually do not take infinitely small time steps we still can hope to obtain a relatively accurate approximation, especially if $\varepsilon$ is small. \textcolor{black}{In particular, if $\varepsilon = C \Delta t = C t_{final}/n$ (where $t_{final}$ denotes the final time and $n$ the number of iterations) we have $(1+\varepsilon)^n = (1+\tfrac{C t_{final}}{n} )^n \leq \exp(C t_{final})$ and thus the scheme is stable, while the error constant is increased by $\exp(C t_{final})$.}
\end{itemize}

@@

%{\color{red}
Finally, we give an illustration of the above comments by running \eqref{vp_linear} ($d=b=1$ and $v_{\max}=3$) 
with a discontinuous initial data to study the impact of high-frequency on the stability of exponential schemes coupled 
with a centered scheme in $v$. 
The initial condition is 
$$
f_0(x, v) = 1 \mbox{ if } \sqrt{(x-\pi)^2+v^2} \leq 1, \mbox{ and } 0  \mbox{ elsewhere}. 
$$
From the stability analysis we know that for each $k$ it holds that $\bar{f}^{n+1}_{k,m} = \phi(z)\bar{f}^{n}_{k,m}$,
where $\phi$ is the stability function. 
Our study enables us to estimate the amplification factor $|\phi(z)|$ by $(1+\varepsilon)$ (uniformly with respect to $k$)  
so that $|\bar{f}^{n+1}_{k,m}|^2 \leq  (1+\varepsilon)^2 |\bar{f}^{n}_{k,m}|^2$. Hence, for a given $\varepsilon$, 
we consider a time step according to Table \ref{tab:ymax_expo} and run 
two exponential schemes, namely Hochbruck-Ostermann and Cox-Matthews,   
for $100$ timesteps. After each time step
we compute $\| f^n\|^2_{\ell^2} / \| f^0\|^2_{\ell^2}$. 
The results can be found in Figure \ref{instab}.   
\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[scale=0.5]{\localPath/figures/error_growup_HO.pdf} & 
\includegraphics[scale=0.5]{\localPath/figures/error_growup_CM.pdf} 
\end{tabular}
\caption{Evolution of $\| f^n\|^2_{\ell^2} / \| f^0\|^2_{\ell^2}$ and of $(1+\varepsilon)^{2n}$ as a function of $n$ for different values of 
$\varepsilon$. Left: Hochbruck-Ostermann scheme. Right: Cox-Matthews scheme. }
\label{instab}
\end{figure}
First, we have a numerical confirmation that $(1+\varepsilon)$ is an estimate of the amplification factor 
(the ratio $\| f^n\|^2_{\ell^2} / \| f^0\|^2_{\ell^2}$  always lies under the curve $n\mapsto (1+\varepsilon)^{2n}$).  
Second, since the number of time steps is fixed, for $\varepsilon=0.1$ the simulation becomes unstable for $n\geq 20$. 
As soon as $\varepsilon$ is small enough, the simulation is stable. 
%As mentioned above, if $\varepsilon$ as $\varepsilon =C \Delta t = Ct_{final}/n$  
%(with $\Delta t=y_{\max} \Delta v$, $y_{\max}$ being given by Table \ref{tab:ymax_expo} and $\Delta v=6/81$), 
%the scheme is  stable and considering a final time $t_{final} = n\Delta t$, the linear analysis leads to 
%\begin{eqnarray*}
%\|{f}^{n}\|_{\ell^2} &=& \|\bar{f}^{n}\|_{\ell^2} \leq  (1+\varepsilon)^n \|\bar{f}^{0}\|_{\ell^2} =  (1+C\Delta t)^{t_{final}/\Delta t} \|\bar{f}^{0}\|_{\ell^2} \nonumber\\
%&\leq & \exp(Ct_{final}) \|\bar{f}^{0}\|_{\ell^2} = \exp(Ct_{final}) \|{f}^{0}\|_{\ell^2}, 
%\end{eqnarray*}
%which gives some practical informations on the stability of the methods since $C$ and $t_{final}$ are known a priori. 
%}

\subsubsection{Linearized WENO5 (LW5) scheme. }

@@

In Figure \ref{cfl_rk44_lw5} (left), we plot the dependence of $\sigma$ with respect to the angle $\varphi\in [-\pi/2, \pi/2]$ 
for Lawson($RK(4, 4)$) coupled with LW5. We also plot (Figure \ref{cfl_rk44_lw5} (right)) the stability domain of Lawson($RK(4, 4)$), 
the eigenvalues for LW5 (normalized by $\Delta v$) and the eigenvalues for LW5 scaled by $\sigma$. 
The CFL number for some Lawson schemes is shown in Table \ref{tab:ymax_weno_Lawson}. It is interesting to note that the time step size is  reduced compared to the centered scheme.
\begin{table}[h]
	\centering
	\begin{tabular}{|c|c|c|c|}
		\hline
		Methods & Lawson($RK(3,2) \; best$) & Lawson($RK(3,3)$) & Lawson($RK(4,4)$) \\
		\hline
		$\sigma $ & $1.344$ & $1.433$   & $1.73$   \\
		\hline  
	\end{tabular}
	\caption{CFL number for some Lawson schemes applied to \eqref{discrete_linear_transport_weno}. }
	\label{tab:ymax_weno_Lawson}
\end{table}

\begin{figure}[h]
  \centering
  \begin{subfigure}[b]{0.4\textwidth}
        \centering \includegraphics[width=\textwidth]{\localPath/figures/cfl_rk44_weno_phi.png}
  \end{subfigure}
  \begin{subfigure}[b]{0.3\textwidth}
        \centering \includegraphics[width=\textwidth]{\localPath/figures/cfl_rk44_weno.png}
  \end{subfigure}
  \caption{Left: $\sigma$ as a function of the angle $\varphi$. Right: stability domain of Lawson($RK(4,4)$) (red), 
  eigenvalues for LW5 normalized by $\Delta v$ (blue) and eigenvalues for LW5 normalized by $\Delta v$ 
  stretched with factor $\sigma=1.73$ (dashed green). } 
  \label{cfl_rk44_lw5}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical simulation: Vlasov-Poisson equations \label{sec:vp}}

>>>>

The goal of this work is to develop high order and efficient numerical methods for nonlinear collisionless kinetic models, such as the Vlasov-Poisson equations or drift-kinetic models. In most situations, the nonlinearity in the transport term originates from the coupling with a Poisson type problem that is used to compute the electric field.

@@

\section{Exponential integrators and Lawson methods\label{sec:expint}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In this paper, we will consider two types of exponential methods. The idea of \textit{exponential integrators} is to use the variation of constants formula to rewrite equation \eqref{eq:expint-eq} in the following form
\[
  u(t_n+\Delta t) = \exp(\Delta t A) u(t_n) + \int_0^{\Delta t} \exp((\Delta t -s)A) F(u(t_n+s)) \dd{s},
\]
where we denote the time step size by $\Delta t >0$ and $t_n = n\Delta t$ with $n\in\mathbb{N}$. This expression is still exact; \ie no approximation has been made. Note, however, that this can not be used as a numerical method as evaluating the integral would require the knowledge of $u(t_n+s)$, which is not available. The idea of an exponential integrator is to approximate the nonlinear part $F(u(t_n+s))$ in terms of the available data. In the simplest case we just evaluate it at the left endpoint. That is, we use $F(u(t_n+s))\approx F(u^n)$. Then we can integrate the term $\exp((\Delta t -s)A)$ exactly and obtain.
\[
  u(t_n+\Delta t) \approx u^{n+1} = \exp(\Delta t A) u^n + \Delta t \varphi_1(\Delta t A) F(u^n),
\]
where $\varphi_1(z)=(\mathrm{e}^z-1)/z$ is an entire function. This is the first order exponential Euler method. In a similar way exponential Runge--Kutta methods can be constructed. We refer to the literature, in particular the review article \cite{Hochbruck:2010}, for more details.

@@

\section{Linear analysis\label{ode}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Lawson methods}
% --------------------------------------------------------------------

@@

\subsection{Exponential integrators}
% --------------------------------------------------------------------

@@

\subsection{Phase space discretization}
% --------------------------------------------------------------------

@@

\paragraph{Centered scheme in $v$\\}
The classic centered scheme is obtained by approximating the velocity derivative in equation \eqref{fourier_x_vlasov} by
$$
  (\partial_v \hat{f}_{k})(v_j) \approx \frac{\hat{f}_{k, j+1}-\hat{f}_{k, j-1}}{2\Delta v},
$$ 
where $\hat{f}_{k,j}$ is an approximation of $\hat{f}_{k}(v_j)$.  Inserting this centered approximation 
in \eqref{fourier_x_vlasov} yields
\begin{equation} \label{eq:half-fourier}
  \partial_t \hat{f}_{k,j} +  i d k\hat{f}_{k,j} +b \frac{\hat{f}_{k,j+1} -\hat{f}_{k,j-1} }{2\Delta v} = 0.
\end{equation}
The system is already diagonal with respect to the index $k$. We now also diagonalize it with respect to the index $j$. To do that we express the function in terms of its Fourier modes with respect to $v$. That is,
\[
  \hat{f}_{k,j} = \sum_m \bar{f}_{k, m}\exp\left(i \frac{2\pi m}{2v_{\max}}  v_j\right),
\]
where $\bar{f}_{k,m}$ denotes the (double) Fourier transform of $f$ with frequency in space $k$ and frequency in velocity $m$. Inserting this into equation \eqref{eq:half-fourier} yields
\begin{equation}
  \label{discrete_linear_transport}
	\partial_t \bar{f}_{k,m} + i dk\bar{f}_{k,m} +b \frac{i\sin(2\pi m \Delta v/(2v_{\max})) }{\Delta v} \bar{f}_{k,m}= 0.  
\end{equation}
We immediately see that this equation is precisely in the form of equation \eqref{ode_linear} as studied in the previous section. We also observe that $\lambda \in i\mathbb{R}$. That is, the eigenvalues for the centered difference approximation lie exclusively on the imaginary axis. One immediate consequence is that for Lawson methods the CFL condition is given by $b \Delta t < C \Delta v$, where $C$ is chosen such that $i [-C,C]$ lies in the domain of stability of the underlying Runge--Kutta method.

\paragraph{Linearized WENO approximation in $v$\\}
A common technique to discretize hyperbolic partial differential equations is to use the so-called weighted essentially non-oscillatory schemes (WENO) schemes. These are nonlinear schemes that limit oscillations in regions where sharp gradients occur, but still yield high order accuracy in smooth regions of the phase space. In the linear case WENO schemes reduce to upwind discretizations. Here, we will consider the LW5 scheme (the linearized version of the WENO5 scheme as considered in \cite{Baldauf:2008, Lunet:2017, Motamed:2010, Wang:2007}) that is given by (from now on we assume w.l.o.g.~that $b>0$)
\begin{align*}
  (\partial_v \hat{f}_k)(v_j) &\approx \frac{1}{\Delta v}\Big(-\frac{1}{30} \hat{f}_{k,j-3} +\frac{1}{4} \hat{f}_{k,j-2} -\hat{f}_{k,j-1} + \frac{1}{3} \hat{f}_{k,j} +\frac{1}{2} \hat{f}_{k,j+1} - \frac{1}{20} \hat{f}_{k,j+2}\Big). 
\end{align*}
We now perform the same analysis as for the centered scheme (see \cite{Baldauf:2008, Crouseilles:2012}). This yields
\begin{align}
  \left( \frac{2\pi }{2 v_{\max}}  i m \bar{f}_{k, m} \approx \right)
  %(i m \bar{f}_{k, m} \approx ) \;\; 
  \mu_m \bar{f}_{k, m} &:= \frac{\bar{f}_{k,m}}{\Delta v}\Big(-\frac{1}{30} e^{-\frac{3i m\pi \Delta v}{v_{\max}}} +\frac{1}{4} e^{-\frac{2i m \pi \Delta v}{v_{\max}}} -e^{-\frac{i m \pi \Delta v}{v_{\max}}}  \nonumber\\
  &\hspace{1cm}+ \frac{1}{3}  +\frac{1}{2} e^{\frac{i m \pi \Delta v}{v_{\max}}} - \frac{1}{20} e^{\frac{2i m \pi \Delta v}{v_{\max}}}\Big). 
  \label{lw5symbol}
\end{align}
We then obtain the equation 
\begin{equation}
	\label{discrete_linear_transport_weno}
	\partial_t \bar{f}_{k,m} + i dk\bar{f}_{k,m} +b \mu_m \bar{f}_{k,m}= 0.
\end{equation}
Once again this is precisely the form of equation \eqref{ode_linear}, where $a=dk\in\mathbb{R}$ and $\lambda=b\mu_m \in\mathbb{C}$. The main difference to the centered difference scheme is that $\lambda$ is not necessarily on the imaginary axis. In fact, the eigenvalues aquire a negative real part which stabilizes the scheme and avoids spurious oscillations, but also adds unphysical dissipation to the numerical method.

\begin{remark}
  Let us remark that performing a Fourier approximation in $v$ is also possible. Using the same notation as before, the counterpart of \eqref{discrete_linear_transport} and \eqref{discrete_linear_transport_weno} in that case is
  $$
    \partial_t \bar{f}_{k,m} + i dk\bar{f}_{k,m} +b i \frac{2\pi m}{2 v_{\max}} \bar{f}_{k,m}= 0.
  $$
  It is worth mentioning that this last equation can be obtained by considering the limit $\Delta v$ goes to zero in  \eqref{discrete_linear_transport} or \eqref{discrete_linear_transport_weno}. The stability condition can be computed as for the centered difference case, since the eigenvalues of the Fourier approximation also lies on the imaginary axis. The CFL condition is given by $b \pi  < C \Delta v$, where $C$ is chosen such that $i [-C,C]$ lies in the domain of stability of the underlying Runge--Kutta method. 
\end{remark}

\subsection{Computing the CFL condition}
% --------------------------------------------------------------------

@@

\subsubsection{Centered scheme in $v$}
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

In the case of centered approximation of the velocity derivative, the Fourier multiplier is a 
pure imaginary complex number (see equation \eqref{discrete_linear_transport}). We thus look for $y_{\max}\in \mathbb{R}_+$ such that the interval $i(-y_{\max},y_{\max}) \subset {\cal D}$, where ${\cal D}$ is the domain of stability for the chosen time integrator.

\paragraph{Lawson integrators\\} 
We simply look for the largest value $y_{\max}$ such that $i(-y_{\max}, y_{\max}) \subset \mathcal{D}$. The corresponding values for a number of schemes are listed in Table \ref{tab:ymax_Lawson}. These values have to be understood in the following way: they induce the CFL condition $b \Delta t\leq y_{\max}\Delta v$ for the discretized equation \eqref{discrete_linear_transport}, where $\Delta t$ denotes the time step size and $\Delta v$ is the velocity mesh size.

@@

\paragraph{Exponential integrators\\}
For the exponential integrators the domain of stability is very sensitive to the value of $(a\Delta t)$. To get an idea of what we can expect, we consider the quantity $$y_{\max}=\min_{(a\Delta t)\in\mathbb{R}} \; y^{exp}_{\max}(a \Delta t).$$ As before, $y^{exp}_{\max}(a\Delta t)$ is the largest value such that : $$i(-y^{exp}_{\max}(a\Delta t), y^{exp}_{\max}(a\Delta t))\subset {\cal D},$$ where ${\cal D}$ is the domain of stability for the chosen exponential time integrator for a given $(a\Delta t)$. Even for relatively simple numerical methods it is not possible to compute this quantity analytically. Thus, we resort to numerical approximations. Unfortunately, it turns out that for most exponential integrators this value is zero. This can be appreciated by considering Figure \ref{fig:expRK_sd} once more. Clearly, there are values of $(a \Delta t)$ such that no relevant part of the imaginary axis (or only half the imaginary axis) is part of the domain of stability. Thus, most exponential integrators are unstable in the von Neumann sense. However, this is not what we observe in practice. In fact, already the results presented in \cite{Crouseilles:2018} indicate that we can successfully run numerical simulations using, for example, the Cox--Matthews scheme. There are two major points to consider here
\begin{itemize}
    \item The $y_{\max}$ obtained is a worst case estimate. In fact, we know that for $\Delta t \to 0$ we regain the stability of the underlying Runge--Kutta method. Thus, for small $(a \Delta t)$ the methods is expected to work well.
    \item As is usually done we have mandated that $\vert \phi(z) \vert \leq 1$. However, strictly speaking this is not necessary for practical simulation. If we assume that $\vert \phi(z) \vert \leq 1+\varepsilon$ and we take $n$ steps the amplification of the error is given by $(1+\varepsilon)^n$. In the limit $n \to +\infty$ this quantity diverges. However, since we usually do not take infinitely small time steps we still can hope to obtain a relatively accurate approximation, especially if $\varepsilon$ is small. In particular, if $\varepsilon = C \Delta t = C t_{final}/n$ (where $t_{final}$ denotes the final time and $n$ the number of iterations) we have $(1+\varepsilon)^n = (1+\tfrac{C t_{final}}{n} )^n \leq \exp(C t_{final})$ and thus the scheme is stable, while the error constant is increased by $\exp(C t_{final})$.
\end{itemize}

@@

Finally, we give an illustration of the above comments by running \eqref{vp_linear} ($d=b=1$ and $v_{\max}=3$) with a discontinuous initial data to study the impact of high-frequency on the stability of exponential schemes coupled with a centered scheme in $v$. The initial condition is 
$$
  f_0(x, v) = 1 \mbox{ if } \sqrt{(x-\pi)^2+v^2} \leq 1, \mbox{ and } 0  \mbox{ elsewhere}. 
$$
From the stability analysis we know that for each $k$ it holds that $\bar{f}^{n+1}_{k,m} = \phi(z)\bar{f}^{n}_{k,m}$, where $\phi$ is the stability function. Our study enables us to estimate the amplification factor $|\phi(z)|$ by $(1+\varepsilon)$ (uniformly with respect to $k$) so that $|\bar{f}^{n+1}_{k,m}|^2 \leq  (1+\varepsilon)^2 |\bar{f}^{n}_{k,m}|^2$. Hence, for a given $\varepsilon$, we consider a time step according to Table \ref{tab:ymax_expo} and run two exponential schemes, namely Hochbruck-Ostermann and Cox-Matthews, for $100$ timesteps. After each time step we compute $\| f^n\|^2_{\ell^2} / \| f^0\|^2_{\ell^2}$. The results can be found in Figure~\ref{instab}.
\begin{figure}
  \centering
  \begin{tabular}{cc}
    \includegraphics[scale=0.5]{\localPath/figures/error_growup_HO.pdf} & 
    \includegraphics[scale=0.5]{\localPath/figures/error_growup_CM.pdf} 
  \end{tabular}
  \caption{Evolution of $\| f^n\|^2_{\ell^2} / \| f^0\|^2_{\ell^2}$ and of $(1+\varepsilon)^{2n}$ as a function of $n$ for different values of $\varepsilon$. Left: Hochbruck-Ostermann scheme. Right: Cox-Matthews scheme.}
  \label{instab}
\end{figure}
First, we have a numerical confirmation that $(1+\varepsilon)$ is an estimate of the amplification factor (the ratio $\| f^n\|^2_{\ell^2} / \| f^0\|^2_{\ell^2}$ always lies under the curve $n\mapsto (1+\varepsilon)^{2n}$). Second, since the number of time steps is fixed, for $\varepsilon=0.1$ the simulation becomes unstable for $n\geq 20$. As soon as $\varepsilon$ is small enough, the simulation is stable. 
%As mentioned above, if $\varepsilon$ as $\varepsilon =C \Delta t = Ct_{final}/n$  
%(with $\Delta t=y_{\max} \Delta v$, $y_{\max}$ being given by Table \ref{tab:ymax_expo} and $\Delta v=6/81$), 
%the scheme is  stable and considering a final time $t_{final} = n\Delta t$, the linear analysis leads to 
%\begin{eqnarray*}
%\|{f}^{n}\|_{\ell^2} &=& \|\bar{f}^{n}\|_{\ell^2} \leq  (1+\varepsilon)^n \|\bar{f}^{0}\|_{\ell^2} =  (1+C\Delta t)^{t_{final}/\Delta t} \|\bar{f}^{0}\|_{\ell^2} \nonumber\\
%&\leq & \exp(Ct_{final}) \|\bar{f}^{0}\|_{\ell^2} = \exp(Ct_{final}) \|{f}^{0}\|_{\ell^2}, 
%\end{eqnarray*}
%which gives some practical informations on the stability of the methods since $C$ and $t_{final}$ are known a priori. 
%

\subsubsection{Linearized WENO5 (LW5) scheme}
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

@@

In Figure \ref{cfl_rk44_lw5} (left), we plot the dependence of $\sigma$ with respect to the angle $\varphi\in [-\pi/2, \pi/2]$ for Lawson($RK(4, 4)$) coupled with LW5. We also plot (Figure \ref{cfl_rk44_lw5} (right)) the stability domain of Lawson($RK(4, 4)$), the eigenvalues for LW5 (normalized by $\Delta v$) and the eigenvalues for LW5 scaled by $\sigma$. The CFL number for some Lawson schemes is shown in Table \ref{tab:ymax_weno_Lawson}. It is interesting to note that the time step size is  reduced compared to the centered scheme.
\begin{table}[h]
	\centering
	\begin{tabular}{|c|c|c|c|}
		\hline
		Methods & Lawson($RK(3,2) \; best$) & Lawson($RK(3,3)$) & Lawson($RK(4,4)$) \\
		\hline
		$\sigma $ & $1.344$ & $1.433$   & $1.73$   \\
		\hline  
	\end{tabular}
	\caption{CFL number for some Lawson schemes applied to \eqref{discrete_linear_transport_weno}.}
	\label{tab:ymax_weno_Lawson}
\end{table}

\begin{figure}[h]
  \centering
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{\localPath/figures/cfl_rk44_weno_phi.png}
  \end{subfigure}
  \begin{subfigure}[b]{0.3\textwidth}
    \centering
    \includegraphics[width=\textwidth]{\localPath/figures/cfl_rk44_weno.png}
  \end{subfigure}
  \caption{Left: $\sigma$ as a function of the angle $\varphi$. Right: stability domain of Lawson($RK(4,4)$) (red), eigenvalues for LW5 normalized by $\Delta v$ (blue) and eigenvalues for LW5 normalized by $\Delta v$ stretched with factor $\sigma=1.73$ (dashed green).} 
  \label{cfl_rk44_lw5}
\end{figure}

\section{Numerical simulation: Vlasov-Poisson equations\label{sec:vp}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
